---
title: "CUNY Data 607 – Week 5 Assignment – Tidying and Transforming Data"
author: "Emilie M Bolduc"
date: "10/1/2017"
output: html_document
---

# Assignment 
1. Create a .CSV file (or optionally, a MySQL database!) that includes all of the information above. You’re encouraged to use a “wide” structure similar to how the information appears above, so that you can practice tidying and transformations as described below.
2. Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy and transform your data.
3. Perform analysis to compare the arrival delays for the two airlines.

# Solution
### Read CSV
Took the data provided and typed it into a CSV table that I then uploaded into R.
```{r}
library(readr)
Airline_delays <- read_csv("~/Documents/CUNY Data 607/data/airlines/Airline_delays.csv")
Airline_delays
```

### Tidy Data
#### 1. Fill first column with complete row names
The carrier names in the first column were not populated across all rows, so needed to fill the remaining rows in that column with the Carrier name.
```{r}
library(tidyr)
Airline <- Airline_delays %>% fill(X1)
Airline
```
Reference: https://blog.rstudio.com/2015/09/13/tidyr-0-3-0/

#### 2. Remove rows with NAs
There was an uncessary empty row in the original CSV file which appeared as NA when uploaded into R, so had to remove it.
```{r}
Airline <- na.omit(Airline)
```

#### 3. Make observations from variables with tidyr 'gather' function
i.e. Change the City column data into observations rows.
```{r}
Airline <- gather(Airline, "City", "n", 3:7)
Airline
```

#### 4. Make variables from observations with tidyr 'spread' function
i.e. Take the "On Time" and "Delayed" row variables and turn them into observations.
```{r}
Airline <- spread(Airline, "X2", "n")
Airline
```

#### 5. Change column names with 'dplyr' fucntions
```{r}
library(dplyr)
Airline <- dplyr::rename(Airline, Carrier = X1)
Airline <- dplyr::rename(Airline, Delayed = delayed)
Airline <- dplyr::rename(Airline, On_Time = 'on time')
Airline
```

I still find the following function quicker and easier...

> colnames(Airline)<- c("Carrier","City","Delayed", "On Time")

### Perform Analysis with 'dplyr'
Taking a look at the data:
```{r}
dplyr::glimpse(Airline)
```

Look at the number of delays by Carrier:
```{r}
select(Airline, Carrier, Delayed)
```

Mean and median of all Delays:
```{r}
dplyr::summarise(Airline, Mean = mean(Delayed), Median = median(Delayed))
```

Look at the number of delays in descending order:
```{r}
dplyr::arrange(Airline, desc(Delayed))
```

Number of delays in ascending order by Carrier:
```{r}
dplyr::arrange(Airline, Carrier, Delayed)
```

The mean and sum of delays by Carrier:
```{r}
Carrier <- Airline %>% group_by(Carrier) %>% 
  summarise(mean = mean(Delayed), sum = sum(Delayed), n = n())
Carrier
```

The mean and sum of delays by City:
```{r}
City <- Airline %>% group_by(City) %>% 
  summarise(mean = mean(Delayed), sum = sum(Delayed)) %>% 
  arrange(desc(mean))
City
```


### Visualize the Analysis
Create table to use for visualizing data:
```{r}
CityCarrier <- Airline %>% group_by(City, Carrier) %>% 
  summarise(mean = mean(Delayed), sum = sum(Delayed))
```

#### Carrier Delays By City
```{r}
library(scales)
library(ggplot2)
LineGraph <- ggplot(CityCarrier, aes(x = City, y = mean))
LineGraph <- LineGraph + geom_line(aes(color=factor(Carrier), group = Carrier))
LineGraph <- LineGraph + scale_color_discrete(name = "Carrier")
LineGraph <- LineGraph + labs(title = "Carrier Delays by City", x = "City", y = "Total Number of Delays")
LineGraph
```
